Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-18 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex

\subsubsection{Relationships between $J$ and $\ell$ for a periodic squared exponential covariance function} \label{subsec_empiric_relations_periodic}

TODO: Is $J$ equal to $m$ or is this something else?

In Appendix~\ref{sec_periodic}, we analyze the accuracy of the linear representation of a periodic squared exponential covariance function and derive the minimum number of terms $J$ in the approximation required to achieve a close approximation to the exact periodic squared exponential kernel as a function of the length-scale $\ell$ of the kernel. Since this is a series expansion of sinusoidal functions, the approximation does not depend on any boundary condition. This relationship between $J$ and $\ell$ is gathered in Figure~\ref{figB1_m_lscale_periodic}.

@@

The numerical equation governing the relationships between $J$ and $\ell$ for a periodic squared exponential covariance function was estimated in eq.~\ref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&J \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{J}
\end{align*}

@@

Furthermore, from Figure~\ref{fig7_lscale_comparison} left it can be seen that using large number of basis functions relative to the wigglyness of the function to be learned (i.e., true length-scale) causes overestimating the underlying functions (see Figure~\ref{fig8_oversmoothed_posterios} as an illustration of the oversmoothing effect as a function of $m$ in a moderate smooth dataset) and, consequently, on the length-scale estimate too (Figure \ref{fig7_lscale_comparison} left). Here it comes out a possible weaknesses of the approach which consists in using too many basis functions relative to the true length-scale causes overestimating the posteriors.

TODO: Please clarify the last sentence above. I don't get that is ment here.

@@

Figure~\ref{fig9_diagnostic} shows the iterative steps of applying the diagnostic procedure explained in Section~\ref{sec_user_guide} over some of the data sets also used in the analysis in Figure~\ref{fig7_lscale_comparison} left. It is clearly visible that by following our recommendations, an optimum solution with minimum computational requirements is achieved in these cases. Figure~\ref{fig10_diagnostic_lscale_comparison} graphically compares the exact GP length-scale and the estimated HSGP length-scale in every iteration and data set. Between two and four iterations, depending on wigglyness of the function to be learned and the distance between the initial guess of the length scale and the true length scale, are sufficient to reach the optimal values of $m$ and $c$. Furthermore, as can be seen in Figure~\ref{fig10_diagnostic_lscale_comparison}, following our diagnostic recommendations avoids overestimation.

TODO: overestimation of what?

TODO: Personally, I think the two examples of the procedure below should go to an appendix, as they take a lot of space and don't add so much additional value.

@@

\begin{figure*}
\centering
\begin{tabular}{ c c c c c c c c c c | c c}
\arrayrulecolor{gray}\hline \\[-3mm]
$\ell_{\text{GP}}$ & iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & $R^2$ & ELPD & RMSE* & \\ 
%\arrayrulecolor{lightgray}\hline \\[-1mm]
%\multirow{5}{*}{ 0.05} 
% & 1 & 0.50 & 1.60 & 6 & 0.127 & FALSE & 0.800 & 0.264 & -0.980 & 0.748 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_1.png}}\\
% & 2 & 0.127 & 1.20 & 17 & 0.061 & FALSE & 0.427 & 0.779 & -0.396 & 0.330 & \\
% & 3 & 0.061 & 1.20 & 35 & 0.059 & TRUE & 0.264 & 0.908 & 0.008 & 0.063 & \\
% & 4 & 0.053 & 1.20 & 40 & 0.059 & TRUE & 0.254 & 0.915 & 0.088 & 0.029 & \\
% \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{5}{*}{ 0.08} 
 & 1 & 0.50 & 1.60 & 6 & 0.169 & FALSE & 0.447 & 0.610 & -0.396 & 0.360 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_2.png}}\\
 & 2 & 0.169 & 1.20 & 13 & 0.068 & FALSE & 0.357 & 0.738 & -0.176 & 0.242 & \\
 & 3 & 0.068 & 1.20 & 31 & 0.080 & TRUE & 0.242 & 0.871 & 0.204 & 0.013 & \\
 & 4 & 0.058 & 1.20 & 36 & 0.081 & TRUE & 0.241 & 0.871 & 0.207 & 0.009 & \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.13} 
 & 1 & 0.50 & 1.60 & 6 & 0.250 & FALSE & 0.461 & 0.625 & -0.482 & 0.330 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_3.png}}\\
 & 2 & 0.250 & 1.20 & 9 & 0.148 & FALSE & 0.352 & 0.780 & -0.135 & 0.153 & \\
 & 3 & 0.148 & 1.20 & 15 & 0.144 & TRUE & 0.321 & 0.807 & -0.074 & 0.084 & \\
 & 4 & 0.105 & 1.20 & 20 & 0.117 & TRUE & 0.297 & 0.830 & -0.006 & 0.013 & \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.25} 
 & 1 & 0.50 & 1.60 & 6 & 0.237 & FALSE & 0.368 & 0.399 & -0.146 & 0.214 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_4.png}}\\
 & 2 & 0.237 & 1.20 & 9 & 0.263 & TRUE & 0.294 & 0.603 & -0.085 & 0.024 & \\
 & 3 & 0.250 & 1.20 & 14 & 0.257 & TRUE & 0.289 & 0.615 & -0.050 & 0.009 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.54} 
 & 1 & 0.50 & 1.60 & 6 & 0.376 & FALSE & 0.272 & 0.858 & 0.034 & 0.013 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_5.png}}\\
 & 2 & 0.376 & 1.20 & 6 & 0.469 & TRUE & 0.277 & 0.855 & 0.029 & 0.051 & \\
 & 3 & 0.239 & 1.50 & 11 & 0.525 & TRUE & 0.272 & 0.858 & 0.036 & 0.014 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.80} 
 & 1 & 1.00 & 3.20 & 6 & 0.516 & FALSE & 0.286 & 0.923 & 0.133 & 0.025 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_6.png}}\\
 & 2 & 0.516 & 1.65 & 6 & 0.975 & TRUE & 0.286 & 0.923 & 0.107 & 0.030 & \\
 & 3 & 0.496 & 3.12 & 11 & 0.772 & TRUE & 0.283 & 0.924 & 0.109 & 0.005 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 1.40} 
 & 1 & 1.00 & 3.20 & 6 & 1.023 & TRUE & 0.283 & 0.561 & 0.139 & 0.010 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_7.png}}\\
 & 2 & 0.521 & 3.27 & 11 & 1.233 & TRUE & 0.281 & 0.566 & 0.136 & 0.003 & \\
 \\
 \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 2.94} 
 & 1 & 0.50 & 1.60 & 6 & 1.192 & TRUE & 0.343 & 0.449 & -0.104 & 0.058 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_8.png}}\\
 & 2 & 0.607 & 3.81 & 11 & 2.577 & TRUE & 0.341 & 0.454 & -0.074 & 0.008 & \\
 \\
 \\
 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{diagnostic on various data sets with varying smoothness. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{fig9_diagnostic}
\end{figure*}

>>>>

\subsubsection{Relationships between $m$ and $\ell$ for a periodic squared exponential covariance function} \label{subsec_empiric_relations_periodic}

In Appendix~\ref{sec_periodic}, we analyze the accuracy of the linear representation of a periodic squared exponential covariance function and derive the minimum number of terms $m$ in the approximation required to achieve a close approximation to the exact periodic squared exponential kernel as a function of the length-scale $\ell$ of the kernel. Since this is a series expansion of sinusoidal functions, the approximation does not depend on any boundary condition. This relationship between $m$ and $\ell$ is gathered in Figure~\ref{figB1_m_lscale_periodic}.

@@

The numerical equation governing the relationships between $m$ and $\ell$ for a periodic squared exponential covariance function was estimated in eq.~\ref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&m \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{m}
\end{align*}

@@

Furthermore, from Figure~\ref{fig7_lscale_comparison} left it can be seen that using large number of basis functions relative to the wigglyness of the function to be learned (i.e., true length-scale) causes overestimating the underlying functions (see Figure~\ref{fig8_oversmoothed_posterios} as an illustration of the oversmoothing effect as a function of $m$ in a moderate smooth dataset) and, consequently, on the length-scale estimate too (Figure \ref{fig7_lscale_comparison} left).

@@

Figure~\ref{fig9_diagnostic} shows the iterative steps of applying the diagnostic procedure explained in Section~\ref{sec_user_guide} over some of the data sets also used in the analysis in Figure~\ref{fig7_lscale_comparison} left. It is clearly visible that by following our recommendations, an optimum solution with minimum computational requirements is achieved in these cases. Figure~\ref{fig10_diagnostic_lscale_comparison} graphically compares the exact GP length-scale and the estimated HSGP length-scale in every iteration and data set. Between two and four iterations, depending on wigglyness of the function to be learned and the distance between the initial guess of the length scale and the true length scale, are sufficient to reach the optimal values of $m$ and $c$. Furthermore, as can be seen in Figure~\ref{fig10_diagnostic_lscale_comparison}, following our diagnostic recommendations avoids the use of a too large number of basis functions relative to the true length scale that would cause overestimating the underlying function and bias the posterior towards smaller length scale values.

TODO: overestimation of what? $\to$ Gabi: Please, see if it's better now.

TODO: Personally, I think the two examples of the procedure below should go to an appendix, as they take a lot of space and don't add so much additional value. $\to$ Gabi: It's fine for me moving them to an appendix. If no-one disagree I'll move them later.

@@

\begin{figure*}
\centering
\begin{tabular}{ c c c c c c c c c c | c c}
\arrayrulecolor{gray}\hline \\[-3mm]
$\ell_{\text{GP}}$ & iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & $R^2$ & ELPD & RMSE* & \\ 
%\arrayrulecolor{lightgray}\hline \\[-1mm]
%\multirow{5}{*}{ 0.05} 
% & 1 & 0.50 & 1.60 & 6 & 0.127 & FALSE & 0.800 & 0.264 & -0.980 & 0.748 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_1.png}}\\
% & 2 & 0.127 & 1.20 & 17 & 0.061 & FALSE & 0.427 & 0.779 & -0.396 & 0.330 & \\
% & 3 & 0.061 & 1.20 & 35 & 0.059 & TRUE & 0.264 & 0.908 & 0.008 & 0.063 & \\
% & 4 & 0.053 & 1.20 & 40 & 0.059 & TRUE & 0.254 & 0.915 & 0.088 & 0.029 & \\
% \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{5}{*}{ 0.08} 
 & 1 & 0.50 & 1.60 & 6 & 0.17 & FALSE & 0.45 & 0.61 & -0.40 & 0.36 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_2.png}}\\
 & 2 & 0.17 & 1.20 & 13 & 0.07 & FALSE & 0.36 & 0.74 & -0.18 & 0.24 & \\
 & 3 & 0.07 & 1.20 & 31 & 0.08 & TRUE & 0.24 & 0.87 & 0.20 & 0.01 & \\
 & 4 & 0.06 & 1.20 & 36 & 0.08 & TRUE & 0.24 & 0.87 & 0.21 & 0.01 & \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.13} 
 & 1 & 0.50 & 1.60 & 6 & 0.25 & FALSE & 0.46 & 0.63 & -0.49 & 0.33 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_3.png}}\\
 & 2 & 0.25 & 1.20 & 9 & 0.15 & FALSE & 0.35 & 0.78 & -0.14 & 0.15 & \\
 & 3 & 0.15 & 1.20 & 15 & 0.15 & TRUE & 0.32 & 0.81 & -0.07 & 0.08 & \\
 & 4 & 0.11 & 1.20 & 20 & 0.12 & TRUE & 0.30 & 0.83 & -0.01 & 0.01 & \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.25} 
 & 1 & 0.50 & 1.60 & 6 & 0.24 & FALSE & 0.37 & 0.40 & -0.15 & 0.21 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_4.png}}\\
 & 2 & 0.24 & 1.20 & 9 & 0.26 & TRUE & 0.30 & 0.60 & -0.09 & 0.02 & \\
 & 3 & 0.25 & 1.20 & 14 & 0.26 & TRUE & 0.29 & 0.62 & -0.05 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.54} 
 & 1 & 0.50 & 1.60 & 6 & 0.38 & FALSE & 0.27 & 0.86 & 0.03 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_5.png}}\\
 & 2 & 0.38 & 1.20 & 6 & 0.47 & TRUE & 0.28 & 0.86 & 0.03 & 0.05 & \\
 & 3 & 0.24 & 1.50 & 11 & 0.53 & TRUE & 0.27 & 0.86 & 0.04 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.80} 
 & 1 & 1.00 & 3.20 & 6 & 0.52 & FALSE & 0.29 & 0.92 & 0.13 & 0.03 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_6.png}}\\
 & 2 & 0.52 & 1.65 & 6 & 0.98 & TRUE & 0.29 & 0.92 & 0.11 & 0.03 & \\
 & 3 & 0.50 & 3.12 & 11 & 0.77 & TRUE & 0.28 & 0.92 & 0.11 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 1.40} 
 & 1 & 1.00 & 3.20 & 6 & 1.02 & TRUE & 0.28 & 0.56 & 0.14 & 0.010 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_7.png}}\\
 & 2 & 0.52 & 3.27 & 11 & 1.23 & TRUE & 0.28 & 0.57 & 0.14 & 0.00 & \\
 \\
 \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 2.94} 
 & 1 & 0.50 & 1.60 & 6 & 1.19 & TRUE & 0.34 & 0.45 & -0.10 & 0.06 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_8.png}}\\
 & 2 & 0.61 & 3.81 & 11 & 2.58 & TRUE & 0.34 & 0.45 & -0.07 & 0.01 & \\
 \\
 \\
 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{diagnostic on various data sets with varying smoothness. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{fig9_diagnostic}
\end{figure*}
